Symbolic regression (SR) fits a symbolic expression to a dataset of expected values.
Amongst its advantages over other techniques are the ability for a practitioner to interpret the resulting expression, determine important features by their usage in the expression, and insights into the behavior of the resulting model (e.g. continuity, derivation, extrema). One of the main algorithms used in SR is genetic programming. The convergence characteristics of SR using GP are still an open issue.
In this work we will use a distributed GP-SR implementation to evaluate the effect of topologies on the convergence of the algorithm.
We introduce and evaluate a tree topology with the aim of finding a new balance between diffusion and concentration.
We use a variation of k-fold cross validation to estimate how accurate a generated solution is in predicting unknown datapoints. This validation technique is implemented in parallel in the algorithm combining both the advantages of the cross validation with the increase in covered search space for each instance. We introduce an incremental approach where the SR tool can start on partial data. This saves the practitioner time as the tool can be used in parallel with the experiment generating the data and allows for a human expert in the loop approach where the partial results of the SR tool can be used to further tune the experiment design. We validate our work on several test problems and a real world use case where a computationally expensive simulator modelling the spread of measles generates the incremental data for our tool.